Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-11 R/SI_Appendix_Datasets.Rmd
revision2: Coherent M&M between Results, M&M, & SI Appendix

@@

>>>>

### Overlap with Hd1 from switchgrass functional validation study

```{r}
df_gam_fl <- read_csv(here("manuscript", "revision2", "SI_Appendix", "Dataset_6_Both_subpopulations_reproductive_growth_date.csv"))
df_gam_fl <- read_csv(here("manuscript", "revision2", "SI_Appendix", "Dataset_4_Gulf_reproductive_growth_date.csv"))
df_gam_fl <- read_csv(here("manuscript", "revision2", "SI_Appendix", "Dataset_5_Midwest_reproductive_growth_date.csv"))

df_gam_fl |> 
  separate(Marker, into = c("Chr", "Pos"), sep = "_", 
           remove = FALSE, convert = TRUE) |> 
  filter(Chr == "Chr04K", between(Pos, 23100000, 23200000)) |> 
  pivot_longer(cols = matches("lfsr"), names_to = "Garden", values_to = "lfsr")

get_significant_results(m_gam_fl_hyp) |> 
  as.data.frame() |> 
  rownames_to_column(var = "Marker") |> 
  separate(Marker, into = c("Chr", "Pos"), sep = "_", 
           remove = FALSE, convert = TRUE) |> 
  filter(Chr == "Chr04K", between(Pos, 23100000, 23200000))

mash_plot_effects(m = m_gam_fl_hyp, i = 13819)
```
Chr04K_23164108 in Both has log10BF ~ 2.3
Chr04K_23179015 in Gulf has log10BF ~ 1.86
Nearby SNPs _not_ significant in Midwest. log10BF ~ 0.1

@@

## Figure S1

Figure 3 is North x Texas GxE effects. Figure S1 is garden x garden GxE effects of individual loci.

Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S1). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

## Figure S2

QTL x Env effects

We conducted quantitative trait loci (QTL) mapping of flowering as functions of five environmental cues that we also used as covariance matrices in mash, and identified eight QTL for flowering date, eight QTL for flowering as a function of day length change two days prior, one QTL for the start of vegetative growth, and two QTL for vegetative growth as a function of daylength change one day prior, all of which showed QTL by environment interactions (SI Appendix, Figure S2).

### 2024-06-11 manuscript/revision2/SI_Appendix/SI_Appendix.qmd
revision2: Coherent M&M between Results, M&M, & SI Appendix

Figures should be cross-referenced like @fig-frog. Each figure should be on its own page. You can control this by placing a pagebreak shortcode, with `{{< pagebreak >}}`.

@@

![Figure](figs/frog.eps){#fig-frog}

@@

| Phenology type | Subpopulation | Effects Selected per GWAS, Strong Subset | Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:---------------|--------------:|-----------------------------------------:|-----------------------------------------:|------------------------------|------------------------------|
| Greenup Date   |          Gulf |                                    19000 |                                    38000 | 45931                        | 91862                        |
|                |       Midwest |                                    24000 |                                    48000 | 49030                        | 98060                        |
|                |          Both |                                    19000 |                                    38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                    19000 |                                    38000 | 50490                        | 100980                       |
|                |       Midwest |                                    33000 |                                    66000 | 63367                        | 126734                       |
|                |          Both |                                    19000 |                                    38000 | 35735                        | 71470                        |

@@

| Greedy Algorithm Iteration \# |     Gulf Greenup Date |           Midwest Greenup Date |    Both Greenup Date | Gulf Flowering Date     | Midwest Flowering Date         | Both Flowering Date     |
|:------------------------------|----------------------:|-------------------------------:|---------------------:|-------------------------|--------------------------------|-------------------------|
| 1                             |              Identity |                       Identity |             Identity | Identity                | Identity                       | Identity                |
| 2                             |         Simple Het. 2 |               Single Effect MO |     Single Effect MI | Simple Het. 1           | Simple Het. 1                  | Single Effect TX3       |
| 3                             |      Single Effect MO |               Single Effect NE |     Single Effect SD | Single Effect MI        | Single Effect MI               | Single Effect TX1       |
| 4                             |      Single Effect NE |              Single Effect TX2 |     Single Effect NE | Single Effect TX3       | Single Effect TX2              | Single Effect OK        |
| 5                             | Daylength (14d prior) |              Single Effect TX1 |    Single Effect TX2 | Single Effect SD        | Single Effect MO               | Single Effect MI        |
| 6                             |      Single Effect MI |               Single Effect MI |     Single Effect OK | Single Effect OK        | Single Effect TX3              | Single Effect NE        |
| 7                             |      Single Effect SD |               Single Effect OK |    Single Effect TX3 | Single Effect NE        | Daylength change (s), 2d prior | Single Effect TX2       |
| 8                             |     Single Effect TX2 |              Single Effect TX3 |        Simple Het. 1 | Single Effect TX1       | Single Effect SD               | Single Effect MO        |
| 9                             |    Cumulative GDD, 6d |                  Simple Het. 1 |    Single Effect TX1 | Single Effect TX2       | \-                             | Single Effect SD        |
| 10                            |     Single Effect TX3 | Daylength change (s), 7d prior |     Single Effect MO | Single Effect MO        | \-                             | Simple Het. 1           |
| 11                            |   Data-driven $PCA_2$ |                             \- |         Temp Ave, 1d | Cumulative rainfall, 7d | \-                             | Cumulative rainfall, 1d |
| 12                            |                    \- |                             \- | Daylength change (s) | \-                      | \-                             | \-                      |

>>>>

Figures should be cross-referenced like @fig-qtl-by-env. Each figure should be on its own page. You can control this by placing a pagebreak shortcode, with `{{< pagebreak >}}`.

@@

![Figure](figs/frog.eps){#fig-garden-by-garden-gxe}

![Figure](figs/frog.eps){#fig-qtl-by-env}

@@

| Phenology type | Subpopulation | Effects Selected per GWAS, Strong Subset | Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:-----------|-----------:|-----------:|-----------:|------------|------------|
| Greenup Date   |          Gulf |                                    19000 |                                    38000 | 45931                        | 91862                        |
|                |       Midwest |                                    24000 |                                    48000 | 49030                        | 98060                        |
|                |          Both |                                    19000 |                                    38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                    19000 |                                    38000 | 50490                        | 100980                       |
|                |       Midwest |                                    33000 |                                    66000 | 63367                        | 126734                       |
|                |          Both |                                    19000 |                                    38000 | 35735                        | 71470                        |

@@

| Greedy Algorithm Iteration \# |     Gulf Greenup Date |           Midwest Greenup Date |    Both Greenup Date | Gulf Flowering Date     | Midwest Flowering Date         | Both Flowering Date     |
|:----------|----------:|----------:|----------:|-----------|-----------|-----------|
| 1                             |              Identity |                       Identity |             Identity | Identity                | Identity                       | Identity                |
| 2                             |         Simple Het. 2 |               Single Effect MO |     Single Effect MI | Simple Het. 1           | Simple Het. 1                  | Single Effect TX3       |
| 3                             |      Single Effect MO |               Single Effect NE |     Single Effect SD | Single Effect MI        | Single Effect MI               | Single Effect TX1       |
| 4                             |      Single Effect NE |              Single Effect TX2 |     Single Effect NE | Single Effect TX3       | Single Effect TX2              | Single Effect OK        |
| 5                             | Daylength (14d prior) |              Single Effect TX1 |    Single Effect TX2 | Single Effect SD        | Single Effect MO               | Single Effect MI        |
| 6                             |      Single Effect MI |               Single Effect MI |     Single Effect OK | Single Effect OK        | Single Effect TX3              | Single Effect NE        |
| 7                             |      Single Effect SD |               Single Effect OK |    Single Effect TX3 | Single Effect NE        | Daylength change (s), 2d prior | Single Effect TX2       |
| 8                             |     Single Effect TX2 |              Single Effect TX3 |        Simple Het. 1 | Single Effect TX1       | Single Effect SD               | Single Effect MO        |
| 9                             |    Cumulative GDD, 6d |                  Simple Het. 1 |    Single Effect TX1 | Single Effect TX2       | \-                             | Single Effect SD        |
| 10                            |     Single Effect TX3 | Daylength change (s), 7d prior |     Single Effect MO | Single Effect MO        | \-                             | Simple Het. 1           |
| 11                            |   Data-driven $PCA_2$ |                             \- |         Temp Ave, 1d | Cumulative rainfall, 7d | \-                             | Cumulative rainfall, 1d |
| 12                            |                    \- |                             \- | Daylength change (s) | \-                      | \-                             | \-                      |

### 2024-06-11 manuscript/revision2/revision2.qmd
revision2: Coherent M&M between Results, M&M, & SI Appendix

Genotypes from the Gulf and Midwest subpopulations had distinct phenological timings and distinct patterns of phenological correlations across our eight common garden sites (@fig-map). At the three Texas common gardens (hereafter 'Texas' gardens), located within the natural range of the Gulf subpopulation, the onset of vegetative growth for Gulf genotypes (hereafter 'green-up') occurred before Midwestern green-up, and the onset of reproductive growth for Gulf genotypes (hereafter 'flowering') occurred after Midwestern flowering (@fig-map A). At the four northernmost common gardens (hereafter 'North' gardens), located within the natural range of the Midwest subpopulation, both green-up and flowering of Gulf genotypes occurred after that of Midwest genotypes. At the Oklahoma common garden, located near the natural range limits of both the Gulf and the Midwest subpopulations, Gulf and Midwest green-up occurred over the same period, and Gulf genotype flowering occurred after Midwestern flowering (@fig-map A). These patterns led to strong negative phenotypic correlations for the onset of vegetative growth between the North and Texas gardens, particularly in the Gulf and across all individuals from the Gulf and Midwest subpopulations (hereafter, 'Both' subpopulations), and contributed to positive phenotypic correlations for the onset of flowering that had larger magnitudes at more northern gardens (@fig-map B).

@@

### Inference of GxWeather effects for vegetative and reproductive timing using *mash*

We next looked for evidence of GxWeather by jointly re-estimating genetic effects of SNPs across all eight common gardens using the *mash* model, which can flexibly capture many modes of covariance across gardens (@fig-map D; SI Appendix, Sections S1-S5, Datasets 1-6). We specified three qualitative categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1; SI Appendix, Section S1). For example, the more similar that the amount of rainfall prior to flowering is for genotypes A and B between gardens in MI and MO, the nearer these covariances will be to one. If rainfall amounts are uncorrelated, covariance will be near zero; if there is rank changing (e.g. high rainfall before flowering of A at MO but not MI, and high rainfall before flowering for B at MI but not MO), this covariance will be near negative one. We specified multiple covariance structures for the canonical, data-driven, and GxWeather categories, each of which represented a specific pattern of covariance in genetic effect size estimates that could be included in the model. The phenotypic correlations for the timing of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf subpopulation and Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, they could be partially or completely captured by the covariance structures specified in the *mash* model.

The GxWeather covariance structures allow hypothesis testing of specific weather variables as cues for the start of vegetative and reproductive growth. Say that a SNP in the gene *Flowering Locus C* (FLC), a well-known flowering time regulator, controls flowering in a photoperiod-dependent manner. In that case, the joint estimate of effects for that SNP could have a high mixture proportion, or mass, on a covariance matrix created using a photoperiod-based environmental cue, such as day length at some interval prior to flowering. In our data, we would infer that the effect of that SNP on flowering was caused by a response to the environmental cue used to construct the GxWeather covariance structure with the largest mass. Overall, loadings of genetic effects on these GxWeather matrices in the model, along with the posterior effect size estimates, provide information on genome-wide patterns of SNP-environment interaction.

We used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood. Five GxWeather covariance structures were selected, zero to two per subpopulation: two for green-up date, and three for flowering date. Of these five, three had mass on them in *mash* models of the strong effects (@fig-covar A-B). Two of these three matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the *mash* model of Gulf green-up fell on a covariance matrix constructed using the covariance of day length 14 days prior to the date of green-up. The covariance matrix for this weather cue was visually similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). *Mash* models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C) for vegetative growth.

For flowering date, distinct GxWeather covariance structures captured covariance in effect sizes for SNPs in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (@fig-covar D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (@fig-covar D), which had negative covariances between Texas and North gardens. Neither of these covariance matrices significantly improved the log-likelihood of the *mash* model of Both subpopulations, and no GxWeather covariance matrices had non-zero mass in models of Both subpopulations (@fig-covar E). In five of the six *mash* models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (@fig-covar C,E); in these five models, the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and single effect matrices with garden-specific effects (SI Appendix, Section S1).

We next characterized the pairwise patterns of effects where we were confident in the sign of the effect at both gardens. If the overall pattern of phenotypic expression differs between a pair of gardens, this difference may be comprised of SNP effects that differ in sign between these gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in the two gardens (@fig-effects A). We used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign.

For green-up date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited a difference in effect sign, between pairs of Texas and North gardens (@fig-effects B; SI Appendix, Fig S1). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects B). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects most often differed in magnitude between gardens within and between the regions (@fig-effects B).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited a difference in effect sign within or between regions (@fig-effects B). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects B), while the majority of effects had no GxE. The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects B). Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S1). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

We sought additional experimental support the occurrence and location of our significant SNP effect re-estimates using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (@fig-qtl A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of five environmental cues that we also used as covariance matrices in *mash*, and identified eight QTL for flowering date, eight QTL for flowering as a function of day length change two days prior, one QTL for the start of vegetative growth, and two QTL for vegetative growth as a function of daylength change one day prior, all of which showed QTL by environment interactions (SI Appendix, Figure S2). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7). All flowering and green-up QTL intervals contained at least one SNP significant in at least one *mash* run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of *mash* SNPs in the 1% tail of significance (the '*mash* 1% tail') within each QTL interval. At the 5% level, three QTL had enrichments of SNPs in the *mash* 1% tail. Overall, there were five significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the *mash* 1% tail in the QTL intervals. Thus, we were able to experimentally support some of our re-estimates of SNP effects with a QTL mapping experiment using a separate mapping population.

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both subpopulations revealed many alleles with sign changes in their effects between the Texas and North gardens (@fig-effects B). As phenological timings are major components of plant fitness, this result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic]; it expands the genetics of local adaptation research to more than two field sites and to a wider range of genetic variation by using effects from GWAS in common gardens located throughout the species' range to determine that rank-changing GxE for phenological timings is common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

@@

Whenever possible, plant material will be shared upon request. Source data and code to replicate these analyses are available at: https://github.com/Alice-MacQueen/pvdiv-phenology-gxe.git. SNP data to replicate these analyses are available from the UT dataverse at https://doi.org/link.

@@

To confirm candidate genomic regions and patterns of allelic effects found in the diversity panel, we analyzed flowering in an outbred pseudo-F2 cross between four individuals, two Midwest and two Gulf individuals. The formation of this mapping population has been described previously [@milano2016genetic]; additional details on QTL mapping can be found in SI Appendix, Section S7. To be directly comparable to the diversity panel data, only 2019 phenology data from the pseudo-F2 cross from the same eight common garden sites were used.

@@

To use *mash* on our diversity panel, we had to specify both a relatively uncorrelated set of covariance matrices, which in our case defined types of GxE and GxWeather between gardens, and we had to specify subsets of SNP effect estimates and standard errors for our traits at each common garden. To specify a set of covariance matrices, we first defined many covariance matrices, including GxWeather matrices that represented the correlation in weather cues between gardenes before the phenological event (SI Appendix, Section S1), then implemented a model selection approach that used a greedy algorithm to evaluate if the log likelihood of the *mash* model was significantly improved as additional covariance matrices were included (SI Appendix, Section S4). To specify subsets of SNP effect estimates and standard errors to use for both the greedy algorithm and with the optimal set of covariance matrices, we first calculated best linear unbiased predictors (BLUPs) for each phenological trait in each genetic subpopulation and each common garden (SI Appendix, Section S2). Next, we determined effect estimates for 8.8 to 12.3 million SNPs per subpopulation by conducting garden-specific GWAS on these BLUPs using $k$ vectors of singular values to correct for population structure, where $k$ was the smallest integer value that made the genomic control coefficient closest to 1 (SI Appendix, Section S2). Singular values were computed using singular value decomposition of the matrix of all SNPs, with iterative SNP pruning and removal of regions in long-range linkage disequilibrium [@privé2017]. Third, to make the *mash* models computationally feasible, we extracted two subsets of SNP effect estimates and standard errors from our GWAS effect estimates: 1) effects from a subset of "strong" tests corresponding to stronger effects on our traits; 2) results from a random subset of all tests to correspond to an unbiased representation of all effects (SI Appendix, Section S3). We used the subset of random effects in our greedy *mash* algorithm (SI Appendix, Section S4) and used the subset of strong effects in our *mash* models with the optimal set of covariance matrices (SI Appendix, Section S5).

>>>>

Genotypes from the Gulf and Midwest subpopulations had distinct phenological timings and distinct patterns of phenological correlations across our eight common garden sites (@fig-map). At the three Texas common gardens (hereafter 'Texas' gardens), located within the natural range of the Gulf subpopulation, the onset of vegetative growth, or 'green-up date', for Gulf genotypes occurred before Midwestern green-up, and the onset of reproductive growth, or 'flowering date', for Gulf genotypes occurred after Midwestern flowering (@fig-map A). At the four northernmost common gardens (hereafter 'North' gardens), located within the natural range of the Midwest subpopulation, both green-up and flowering of Gulf genotypes occurred after that of Midwest genotypes. At the Oklahoma common garden, located near the natural range limits of both the Gulf and the Midwest subpopulations, Gulf and Midwest green-up occurred over the same period, and Gulf genotype flowering occurred after Midwestern flowering (@fig-map A). These patterns led to strong negative phenotypic correlations for the onset of vegetative growth between the North and Texas gardens, particularly in the Gulf and across all individuals from the Gulf and Midwest subpopulations (hereafter, 'Both' subpopulations), and contributed to positive phenotypic correlations for the onset of flowering that had larger magnitudes at more northern gardens (@fig-map B).

@@

### Inference of genome-wide patterns of GxE and GxWeather for vegetative and reproductive timing

We expected patterns of GxE and GxWeather to vary at the locus, subpopulation, and trait level; thus we were interested in modeling the different ways effects might covary across our eight common gardens. We jointly re-estimated genetic effects of a subset of SNPs across all eight common gardens using *mash* models, which can flexibly capture many modes of covariance across gardens (@fig-map D; SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects, we used subsets of effects from garden- and subpopulation GWAS (SI Appendix, Section S2-S3; Table S1). To obtain covariance structures, we specified both covariance structures introduced in the initial *mash* manuscript, and GxWeather covariance matrices estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1; SI Appendix, Section S1). Then, we used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood (SI Appendix, Section S4; Table S2). Finally, we fit *mash* models using the selected set of covariance matrices on a subset relatively unlinked (r^2^ \< 0.2), 'strong' genetic effects with low *p*-values (SI Appendix, Section S5).

The loadings of genetic effects onto the GxE and GxWeather covariance matrices included in each model provide information on genome-wide patterns of SNP-environment interaction. For example, the phenotypic correlations for the onset of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf subpopulation and Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, they could be partially or completely captured by the covariance structures specified in the *mash* model. In this case, the joint estimate of effects for many SNPs could have high mixture proportions, or mass, on a covariance matrix that captures the negative correlation between Texas and North gardens. In our data, daylength 14 days prior to green-up date has negative correlations between Texas and North gardens (@fig-covar A); if many SNP effects have mass on this GxWeather matrix, we would infer that the effects of those SNPs have that form of GxWeather covariation.

Eight GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation: five for green-up date, and three for flowering date (SI Appendix, Table S2). Of these eight, five had mass on them in *mash* models of the strong effects (@fig-covar A-B). Three of these five matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while two had all positive or near-zero covariances.

For green-up date, SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the *mash* model of Gulf green-up fell on a covariance matrix constructed using the covariance of day length 14 days prior to the date of green-up. The covariance matrix for this weather cue was visually similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). *Mash* models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C) for vegetative growth.

For flowering date, distinct GxWeather covariance structures captured covariance in effect sizes for SNPs in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (@fig-covar D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (@fig-covar D), which had negative covariances between Texas and North gardens. Neither of these covariance matrices significantly improved the log-likelihood of the *mash* model of Both subpopulations, and no GxWeather covariance matrices had non-zero mass in models of Both subpopulations (@fig-covar E). In five of the six *mash* models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (@fig-covar C,E); in these five models, the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and single effect matrices with garden-specific effects.

### Frequency of rank-changing GxE in significant SNP effects

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (@fig-effects A).

For green-up date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited rank-changing GxE, or a difference in sign between pairs of Texas and North gardens (@fig-effects B; SI Appendix, Fig S1). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects B). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects most often differed in magnitude between gardens within and between the regions (@fig-effects B).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited rank-changing GxE, or a difference in effect sign within or between regions (@fig-effects B). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects B), while the majority of effects had no GxE. The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects B). Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S1). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

We sought additional experimental support the occurrence and location of our significant SNP effect re-estimates using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (@fig-qtl A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of five environmental cues that we also used as covariance matrices in *mash*, and identified eight QTL for flowering date, eight QTL for flowering as a function of day length change two days prior, one QTL for the start of vegetative growth, and two QTL for vegetative growth as a function of daylength change one day prior, all of which showed QTL by environment interactions (SI Appendix, Figure S2). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7); the QTL on Chr04K overlapped a gene recently functionally validated for flowering in switchgrass [@choi2023]. All flowering and green-up QTL intervals contained at least one SNP significant in at least one *mash* run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of *mash* SNPs in the 1% tail of significance (the '*mash* 1% tail') within each QTL interval. At the 5% level, three QTL had enrichments of SNPs in the *mash* 1% tail. Overall, there were five significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the *mash* 1% tail in the QTL intervals. Thus, we were able to experimentally support the genomic windows of some re-estimates of significant SNP effects with a QTL mapping experiment using a separate mapping population.

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both subpopulations revealed many alleles with rank-changing GxE, or sign changes in their effects between the Texas and North gardens (@fig-effects B). As phenological timings are major components of plant fitness, this result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic]. In addition, it expands the genetics of local adaptation research to more than two field sites and to a wider range of genetic variation by re-estimating effects from GWAS to determine that rank-changing GxE for phenological timings is common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

@@

Whenever possible, plant material will be shared upon request. Source data and code to replicate these analyses are available at: https://github.com/Alice-MacQueen/pvdiv-phenology-gxe.git. SNP data to replicate these analyses are available from the UT dataverse at <https://doi.org/10.18738/T8/A604BU>.

@@

To confirm candidate genomic regions found using *mash* on the diversity panel, we analyzed flowering in an outbred pseudo-F2 cross between four individuals, two Midwest and two Gulf individuals. The formation of this mapping population has been described previously [@milano2016genetic]; additional details on QTL mapping can be found in SI Appendix, Section S7. To be directly comparable to the diversity panel data, only 2019 phenology data from the pseudo-F2 cross from the same eight common garden sites were used.

@@

To use *mash* on our diversity panel, we had to specify both a relatively uncorrelated set of covariance matrices, which in our case defined types of GxE and GxWeather between gardens, and we had to specify subsets of SNP effect estimates and standard errors for our traits at each common garden. To specify a set of covariance matrices, we first defined many covariance matrices, including GxWeather matrices that represented the correlation in weather cues between gardenes before the phenological event (SI Appendix, Section S1), then implemented a model selection approach that used a greedy algorithm to evaluate if the log likelihood of the *mash* model was significantly improved as additional covariance matrices were included (SI Appendix, Section S4). To specify subsets of SNP effect estimates and standard errors to use for both the greedy algorithm and with the optimal set of covariance matrices, we first calculated best linear unbiased predictors (BLUPs) for each phenological trait in each genetic subpopulation and each common garden (SI Appendix, Section S2). Next, we determined effect estimates for 8.8 to 12.3 million SNPs per subpopulation by conducting garden-specific GWAS on these BLUPs using $k$ vectors of singular values to correct for population structure, where $k$ was the smallest integer value that made the genomic control coefficient closest to 1 (SI Appendix, Section S2). Singular values were computed using singular value decomposition of the matrix of all SNPs, with iterative SNP pruning and removal of regions in long-range linkage disequilibrium [@privé2017]. Third, to make the *mash* models computationally feasible, we extracted two subsets of SNP effect estimates and standard errors from our GWAS effect estimates: (i) effects from a subset of "strong" tests corresponding to stronger effects on our traits; (ii) results from a random subset of all tests to correspond to an unbiased representation of all effects (SI Appendix, Section S3). We used the subset of random effects in our greedy *mash* algorithm (SI Appendix, Section S4) and used the subset of strong effects in our *mash* models with the optimal set of covariance matrices (SI Appendix, Section S5).
